Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-22 paper.tex
Ana's updates

\subsection{Recombination breakpoint intervals}
\label{sec:breakpoint_intervals}
It is rarely possible to be precise about the position on the genome at which
a recombinant sequence switches from inheriting from one parent to another.
Even if we observe the recombinant sequence before signficant
divergence occurs (during onward transmission), if the two parent sequences
are similar there is no way to distinguish them. Here we define the
interval for a particular breakpoint as the genome coordinates over which the
sequences for the left and right parent nodes are identical. The right-hand
extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
(Section~\ref{sec:ls}) and the left bound is then derived by
sequence comparison of the parents.

@@

While the detailed interpretation of Figure~\ref{fig:breakpoint-distribution}
requires further study, a few key properties stand out.[TODO, write this
better. ] Firstly, in several cases there is a clear relationship between
gene boundaries and the breakpoint intervals. For example, ORF7a and OR7b
show sharp differences in the number of intersecting intervals within
these genes. This could be due to a number of factors, such as [BIOLOGY]
or the presence of sites with artefactually high numbers of mutations
in the intergenic regions (see the next paragraph for more discussion).
Secondly, we see substantially fewer intervals intersecting with the N
gene than other regions, in agreement with the results of
\cite{Turakhia2022-it}.

It is important to note that the exact endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents. The presence of problematic sites that are enriched
for errors will therefore tend to artificially truncate breakpoint intervals
at these sites.
It is therefore helpful to compare the number of interecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For example, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and corresponds to the
left endpoint of 9 intervals and the right of 57.
The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
at this site are a likely indicator that this
homopolymeric region is prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
% There can be good biological reasons for this too though
However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
have the same effect of truncating intervals. At this site we have
770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
of polymorphism maintained by selection [FIXME!!], and 58 intervals
with endpoints within one base of this site.

>>>>

\subsection{Recombination breakpoint intervals}
\label{sec:breakpoint_intervals}
It is rarely possible to be precise about the position on the genome at which
a recombinant sequence switches from inheriting from one parent to another.
Even if we observe the recombinant sequence before signficant
divergence occurs (during onward transmission), if the two parent sequences
are similar there is no way to 
% distinguish them. 
accurately specify the breakpoint.
Here we define the
interval around a particular breakpoint as the genome coordinates over which the
sequences for the left and right parent nodes are identical. The right-hand
extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
(Section~\ref{sec:ls}), and the left endpoint is then derived by directly
comparing the parent sequences.

@@

Emphasising that further work is required to filter out spuriously inferred
recombination events and improve the accuracy of the inferred breakpoint
intervals, we can draw two preliminary conclusions from 
Figure~\ref{fig:breakpoint-distribution}.
% While the detailed interpretation of Figure~\ref{fig:breakpoint-distribution}
% requires further study, a few key properties stand out.[TODO, write this
% better. ] 
Firstly, in several cases there is a clear relationship between
gene boundaries and the breakpoint intervals. For example, ORF7a and OR7b
show sharp differences in the number of intersecting intervals within
these genes. This could be due to a number of factors, such as [BIOLOGY]
or the presence of sites with artefactually high numbers of mutations
in the intergenic regions (see the next paragraph for more discussion).
Secondly, we see substantially fewer intervals intersecting with the N
gene than other regions, in agreement with the results of
\cite{Turakhia2022-it}.

It is important to note that the precise endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents. 
% The presence of problematic sites that are enriched
% for errors will therefore tend to artificially truncate breakpoint intervals
% at these sites.
% It is therefore helpful to compare the number of interecting intervals
% with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
% For example, position 28,271 (located in the ORF8-N intergenic
% region) has the largest number of mutations and corresponds to the
% left endpoint of 9 intervals and the right of 57.
% The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
% at this site are a likely indicator that this
% homopolymeric region is prone to sequencing errors
% and should be included in the list of ``problematic sites'' that
% are excluded from analysis (Section~\ref{sec:data_preprocessing}).
% There can be good biological reasons for this too though
% However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
% have the same effect of truncating intervals. At this site we have
% 770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
% of polymorphism maintained by selection [FIXME!!], and 58 intervals
% with endpoints within one base of this site.
Thus, breakpoint intervals will tend to be truncated at sites which are
hypermutable, either due to being prone to sequencing errors or
for biological reasons such as selection pressure.
It is therefore helpful to compare the number of intersecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For instance, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and appears as an endpoint
of 66 intervals.
The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
On the other hand, 58 of the breakpoint intervals have endpoints
within one base of site 27,972, which has undergone 770 mutation events 
(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
truncating ORF8, and it has been posited that the truncation is neutral
or advantageous for transmission, and disadvantageous within-host 
\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation 
at this site is due to selection.
